---
title: "Expression Set Class Adapted from Falcon, Morgan and Gentleman 2007"
author: Ella Deyett
date: March 27, 2020
output: 	
	html_document:
		toc: true
		toc_depth:2 
		toc_float: 
		  collapsed:false
		  smooth_scroll:false
		highlight: tango
---

#Introdcution
	ExpressionSet consists of and coordinates between many different components including:
		assayData: hint at the methods used to access different data components 
		meta-data: describing samples in the experiment 
		phenoData: annotations and metadata about the samples 
		featureData/annotation: protocol used for processing each sample 
		experimentData:
		protocolData:

# Preliminaries 
##Install/Load packages
```{r}
BiocManager::install("Biobase")
library("Biobase")
packageVersion("Biobase")
```

##Building an ExpressionSet from .CEL and other files and from Scratch
	CEL files crom from chip manufacturer hardward and R packages can read this data and make it into an expressionset 
			Can be converted using: as(object, "ExpressionSet")

###Assay Data
	F(eatures)= Row
	S(amples)=Columns
	Usually this is in a tab delimited text files so can read in with read.table and then convert to a matrix
```{r}
dataDirectory<-system.file("extdata", package="Biobase")
exprsFile<-file.path(dataDirectory, "exprsData.txt")
exprs<-as.matrix(read.table(exprsFile, header=TRUE, sep="\t", row.names=1, as.is=TRUE))

#Data assessment
class(exprs)
dim(exprs)
colnames(exprs)
head(exprs)

mimimalSet<-ExpressionSet(assayData=exprs)
```

###Phenotypic Data
	Summarizes information about samples (condition, tissue, age, treatment etc)
		S(amples)= rows
		V(variables)=columns

```{r}
pDataFile<-file.path(dataDirectory, "pData.txt")
pData <- read.table(pDataFile, row.names=1, header=TRUE, sep="\t")

dim(pData)
rownames(pData)
#Should match colnames of assay data!
summary(pData)


#Test if they match 
all(rownames(pData)==colnames(exprs))
names(pData)
sapply(pData, class)
pData[c(15,20),]
pData[pData$score>0.8,]

#To better explain what the variables mean we can create a metadata table
metadata<-data.frame(labelDescription=c("Patient Gender", "Case/Control status", 
			"Tumor progress on XYZ scale"), row.names=c("gender", "type", "score"))
head(metadata)


#AnnotatedDataFrame class in biobase connects our phenotypic data for our metadata

phenoData<-new("AnnotatedDataFrame", data=pData, varMetadata=metadata)

phenoData
head(pData(phenoData))

phenoData[c("A","Z"), "gender"]

pData(phenoData[phenoData$score>0.8,])

```

##Annotations and feature data
	
```{r}
annotation<-"hgu95av2"

#Create an object describing the experimental design as a MIAME object

experimentData<-new("MIAME", 
	name="Pierre Fermat", 
	lab="Francis Galton Lab", 
	contact="pfermat@lab.not.exist", 
	title="Smoking-Cancer Experiment", 
	abstract="An example ExpressionSet", 
	url="www.lab.not.exist", 
	other=list(notes="Created from text files"))


```


##Putting it all Together
```{r}
exampleSet<-ExpressionSet(assayData=exprs, 
		phenoData=phenoData,
		experimentData=experimentData,
		annotation='hgu95av2')


```

#The basics of ExpressionSets
```{r}
#Summary of expression set
exampleSet

#Extracting data from expression set
exampleSet$gender[1:5]
exampleSet$gender[1:5]=="Female"


#Retrieve Features names
featureNames(exampleSet)[1:5]

#Retrieve Sample Names
sampleNames(exampleSet)[1:5]

#Retrieve Var Labels
varLabels(exampleSet)


#Extract the expression matrix and the annotated data from 

mat<-exprs(exampleSet)
dim(mat)

adf<-phenoData(exampleSet)
adf


```

##Subsetting
	So important and useful! You want a particular piece of data from the expression set but also want it to still be connnected to the phenotypic data! 

	Keep everything together
```{r}
vv<-exampleSet[1:5, 1:3]
dim(vv)
featureNames(vv)
sampleNames(vv)

#What about subsetting to just the males
males<-exampleSet[,exampleSet$gender=="Male"]
males
exampleSet

```

#All Good code ends with package info!
```{r}
sessionInfo()

```
